\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-03 sections/evaluation.tex
Wide images
Recent changes

Over 12 months, several studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped in 3 phases, each tested a different version of \chameleon{} with human participants. The number of participants and their haskell experience is shown in the table below.

A summary of the timeline of the user studies, their major questions and conclusions is given in Figure~\ref{fig:timeline}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/timeline.pdf}
    \caption{The timeline of \chameleon{}  development and 
    evaluation.}
    \label{fig:timeline}
\end{figure}

@@

Participant recruiting channel was the social news aggregation website Reddit. Specifically, our user study advertisements were posted in the Haskell programming language community "r/haskell" and a general programming language-focused community "r/programminglanguages". Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. The participation is fully anonymized. The detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

One consequence of our recruiting approach is it is harder for previous participants to enter an later study. To minimise the lurking variable of previous experience, we use new code challenge every study and conduct trial runs in every study to bring new participants up to speed.

\subsubsection*{\textbf{Experiment setting}}
The experiment took place remotely and unsupervised. Participants took the study online via web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups. The downside is that to intervene when a user encounters bugs or usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on every function of the tool interface and interactive features. Participants were also shown the important functionalities of the tool interface in a summarized cheat sheet. Participants had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection starts.

All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each user study, participants were required to complete a series of programming tasks (8 for user studies 1 - 3, 9 for user study 4). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to make the code type check with the help of the tool given for this task.

% During each study, a participant received two (user studies 1 - 3) or three (user study 4) different debugging tools. The debugging tools were different systems (user studies 1 - 3) or different feature sets of the same system (user study 4). For each participant, the debugging tools alternative through consecutive tasks to ensure the fatigue level does not impair the rigorousness of data collection.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three attempts or spent over 1 minute on the task.

After completing all the tasks (8 in user studies 1-3, 9 in user study 4), users are prompted to fill in a debriefing survey. The survey questions include their Haskell programming experience and feedback on the tools and feature sets participants used during the study.

We used an browser session recording tool~\cite{openreplay} to record the user study sessions. These recordings are used to find usability issues in the study and recognize general patterns. However, the recording technology does not provide high enough sampling rate for us to be confidently used for rigorous analysis.

\subsection{\chameleon{} User Studies}

\subsubsection{\textbf{Version 1}}  
\chameleon{} version 1 provides the rewrite of the Chameleon type inference engine and the type compare tool to show the alternative types and possible error locations in matching colors.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
    \caption{
        \chameleon{} Version 1
        }
    \label{fig:chameleon-v1}
\end{figure}

Two user studies were conducted to test \chameleon{} version 1, to compare the effectiveness of solving type errors using \chameleon{} and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type error in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim.

Eight tasks were given in both studies. In the first study, the tasks were taken from a exercises from the exercises of the Haskell programming units. These tasks cover a variety of type errors. \tf{break down the type error classes.} However these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options. In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{githubHaskell}. The authors manually added type errors.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
    \caption{}
    \label{fig:study-process}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
    \caption{Time to complete by task in user study 1a with confidence interval}
    \label{fig:r1-analysis}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
    \caption{Time to complete by task in user study 1b confidence interval}
    \label{fig:r2-analysis}
\end{figure}

@@

From the data collected during user study 1 (\ref{fig:r1-analysis}), we could not identify which group performed better (p-value =0.2041, Wilcoxin signed test). For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. 
The most common feedback (7 out of 35) from the user study 1 was that the tasks were too trivial to invite meaningful evaluation. One participant put it, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice".

One interesting exception is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8:
\begin{itemize}
    \item {It has a longer source file than other tasks (only shorter than task 6, but task
    6 contains two independent type errors while task 8 is one connected task
    error);}
    \item {It is more complex than others (involves abstract data types and function application); and
    }
    \item {
        GHC struggles to produce a relevant error message for this type of error.
    }
  \end{itemize}
It can be further shown by examining the process some participants took to solve the problem.

@@

In comparison, the \chameleon{} group solved the type error faster than the GHC group in almost all tasks (p =0.00742, Wilcoxin signed test) in study 2 (figure \ref{fig:r2-analysis}), barring task 1. The authors are unclear about the cause of low performance from \chameleon{} in task 1. It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. From the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.

In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "Its most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgements. This makes it quite easy to figure out how to rectify errors." reported another participant.

\subsubsection{\textbf{Version 2}}  \label{sub:us4}

Version 2 is latest version of \chameleon{}. Not only the added features the deduction steps and candidate expressions are We designed an exploratory study to get better understanding of how users learn and use the multi-mode design of \chameleon{} version 3. During the study the initial mode of each task alternative through the three different modes, and repeat three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants are free to switch to other modes at any time.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
    \caption{}
    \label{fig:study-process}
\end{figure}

@@

The study of \chameleon{} 2 is exploratory, in that we were hoping to let programmers discover their own way of using the tool. In post-hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. We share a few of our observations in this section.

The most striking feature of the data is users' tendency of using tools differ wildly. That is to say, some users used the feature extensively however some users completed the tasks without actively explore the given information. Based on this difference we divided the users into three groups.

@@

    \hline
        Interaction level & Description \\ \hline
        Minimal Interaction & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        High Interaction & Users did everything from the Common feature group but used features specific to Balanced and Advanced mode, such as activating steps and expression cards. \\ \hline

\end{tabularx}

As shown in  figure \ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better in the tasks, and the lowest level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As discussed in user studies 1 and 2, it is likely related to the nature of these tasks. Tasks 4 and 6 are shorter tasks in the study. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less important. On the other hand, task 9 is the longest task of all. It also involves harder problems such as mutually recursive type definitions.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-2.pdf}
    \caption{Time to complete by task in user study 2 with 95\% confidence interval}
    \label{fig:r4-analysis}
\end{figure}

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/r4-task9.png}
%     \caption{
% High interaction users were able to locate the exact line for a potential fix after navigating to deduction step 6 or 7, which directly revealed the true cause of the type error. After this, 12 out of 15 high interaction users provided the most ideal fix in the first try. Low interaction group  takes longer time to examine the problem, provided wrong fixes in the initial trials. One participant fixed a minor run-time issue (a problematic pattern matching of `[attributes]` on line 18) however failed to identify the culprit of the type error.
%     }
%     \label{fig:r4-task9}
% \end{figure}

Another observation is when using the mode switching feature of \chameleon{}, programmers generally switching form a less informative display to a more informative one (Fig. \ref{fig:r4-mode-switching}).

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/mode-switching.pdf}
    \caption{
        The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

>>>>

Three studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants. The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{images/timeline.pdf}
    \caption{The timeline of \chameleon{}  evaluation.}
    \label{fig:timeline}
\end{figure*}

@@

The recruiting channel we used was the social news aggregation website Reddit. Specifically, our user study advertisements were posted in the Haskell programming language community \textit{r/haskell} and a general "programming language" focused community \textit{r/programminglanguages}. Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. The participation is fully anonymized. The detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

One consequence of our recruiting approach is it is harder for previous participants to enter a later study. To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.

\subsubsection*{\textbf{Experiment setting}}
The experiments took place remotely and unsupervised. Participants took the study online via a web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups. The downside is that intervening when users encounter usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface. Participants had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started.

All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to make the code type check with the help of the given tool.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.

After completing all the tasks, users are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording"  to find usability issues in the study and recognize general patterns. The only reason this format is inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.

One other use of the session recording is for outlier trimming.   identify participants who left the computer for a long period of time. We first ran a statistical analysis to identify the long gap (2sd) between program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

\subsection{\chameleon{} Human Studies}

\subsubsection{\textbf{\chameleon{} 1}}  
\chameleon{} 1 is an earlier iteration of \chameleon{} development. This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features. Key features in \chameleon{} 1 include showing two alternative types for the ill-typed expression,  providing all possible locations that contribute to the type error, and dividing possible locations into two groups based on which alternative type can be inferred. In terms of features, \chameleon{} 1 is equivalent to \chameleon{} 2 in basic mode.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
%     \caption{
%         \chameleon{} 1
%         }
%     \label{fig:chameleon-v1}
% \end{figure}

Two user studies (1a & 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. The procedure of these studies is shown in figure \ref{fig:procedure-1}.

Eight tasks were given in both studies. In study 1a, the tasks were taken from exercises from the Haskell programming units. These tasks cover a variety of type errors. However, these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options. In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}. The authors manually added type errors.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
    \caption{}
    \label{fig:procedure-1}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
    \caption{Time to complete each task in user study 1a with 95\% confidence interval}
    \label{fig:analysis-1a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
    \caption{Time to complete each task in user study 1b with 95\% confidence interval}
    \label{fig:analysis-1b}
\end{figure}

@@

From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better. For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. The most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation. One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8. It has a lengthier source file and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error.

In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. The authors are unclear about the cause of low performance from \chameleon{} in task 1. It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training. For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.

In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

% \begin{itemize}
%     \item {It has a longer source file than other tasks (only shorter than task 6, but task
%     6 contains two independent type errors while task 8 is one connected task
%     error);}
%     \item {It is more complex than others (involves abstract data types and function application); and
%     }
%     \item {
%         GHC struggles to produce a relevant error message for this type of error.
%     }
%   \end{itemize}

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}

\chameleon{} 2 is latest version of \chameleon{}. Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default as well. A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 was designed to evaluate \chameleon{} 2.  We were trying to answer the research question: \textit{How do programmers use the interactive features in \chameleon{} 2?}. More specifically, we were trying to inquiry the following three sub-questions:
\begin{itemize}
    \item Do programmers prefer switching modes during debugging type errors?
    \item What are programmers' preferences among the three modes provided by \chameleon{} 2?
    \item How do programmers use the advanced features provided by \chameleon{} 2?

\end{itemize}

During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants can switch to other modes at any time (figure \ref{fig:procedure-2}).

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
    \caption{}
    \label{fig:procedure-2}
\end{figure}

@@

The study of \chameleon{} 2 is an exploratory study. We were expecting to let programmers discover their own way of using the tool. In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. We share a few of our observations in this section.

The most striking feature of the data is users' tendencies to use the tool differ wildly. Some users used the feature extensively however some users completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups.

@@

    \hline
        Interaction level & Description \\ \hline
        Minimal Interaction & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        High Interaction & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}

As shown in  figure \ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better in the tasks, and the lowest level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As discussed in user studies 1 and 2, it is likely related to the nature of these tasks. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-2.pdf}
    \caption{Time to complete each task in study 2 with 95\% confidence interval}
    \label{fig:r4-analysis}
\end{figure}

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/r4-task9.png}
%     \caption{
% High interaction users were able to locate the exact line for a potential fix after navigating to deduction step 6 or 7, which directly revealed the true cause of the type error. After this, 12 out of 15 high interaction users provided the most ideal fix in the first try. Low interaction group  takes longer time to examine the problem, provided wrong fixes in the initial trials. One participant fixed a minor run-time issue (a problematic pattern matching of `[attributes]` on line 18) however failed to identify the culprit of the type error. This location is not pinpointed by \chameleon{} and therefore it is understandable that programmers who used the deduction steps would be able to ignore this part of the source code.
%     }
%     \label{fig:r4-task9}
% \end{figure}

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig. \ref{fig:r4-mode-switching}). 
This observation suggests two characteristics of using multi-mode debugging tools. First, programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/mode-switching.pdf}
    \caption{
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

### 2022-12-03 sections/introduction.tex
Wide images
Recent changes

Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer2012static} and codebase maintainability~\cite{kleinschmager2012static}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley2019next700}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{so2022survey}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

Programmers find challenges adopting typed languages in their practice. They tend to be verbose, hard to learn, and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau2011measuring, tirronen2015understanding}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{yang2000improved, some more proof}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues poses challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool to visualize critical context of a type error (where it happened or could have happened, and which chunks of code caused it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature , and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{chameleon} and desktop versions. The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak2007history}, and now it can be found in most popular languages such as C\#~\cite{csharpgenerics}, Java~\cite{javagenerics}, and TypeScript~\cite{tsgenerics}.

@@

This paper makes the following contributions:
\begin{itemize}
\item { We provide the design and implementation of the \chameleon{}, to visualize crucial context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  }
\item {
    We report the result of four experiments designed to evaluate \chameleon{}.}
\end{itemize}

>>>>

Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues pose challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions. The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

@@

This paper makes the following contributions:
\begin{itemize}
\item { We provide the design and implementation of the \chameleon{}, to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  }
\item {
    We report the result of four experiments designed to evaluate \chameleon{}.}
\end{itemize}

### 2022-12-03 sections/related-work.tex
Recent changes

\section{Related Work}

Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process. Haack and Wells~\cite{haack2004type} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Researchers have proposed solutions to improve the type inference process. Type error slicing~\cite{haack2004type} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and MUS manipulation are used to generate these slices. The language supported in Haack's work was a subset of standard ML. The original Chameleon~\cite{stuckey2003interactive} used a similar technique but extended it to support advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface (Fig.~\ref{fig:original-chameleon}). Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.

Another direction to approach type error debugging is suggesting fixes. Sheng Chen's works [1,2] on counter factual typing is able to suggesting changes to make the ill-typed program typecheck. While this is promising ground, the work is not evaluated with human participants.  \cite{lerner2007searchingtypeerror} used an approach by repeatedly modifying the abstract syntax tree and querying the compiler/type checker to know whether the program still well-typed. This approach allows the algorithm to produce suggestions the syntax changes that could fix the type error. Benefiting from not needing a type system impelemntation and being language agnositic, the underlying idea was implemented and evaluated on 10 CS students. The system can be successfully adopted in not suitable for teaching and learning purpose, as it is not able to explain why the type error happened and why fixes are suggested. The lack of explanability is because the information was witheld by the type checker and the inference algorthm does not know (need to know) how types are computed.

One related approach is CEME (compiler error message enhancement). Prather showed that ECEM casts a positive result in understanding the type error through a series of mixed methods study. Decafe is a tool that is able to rephase the compier error message into an enhanced version. A study of over 200 CS1 students showed that CEME reduce overall errors in their assignments. Berik proposed a framework of constructing compiler error message based on the augumentation theory. In this proposal,  Berik argued that error messages  follow a simple argumentation layout or an exntended argumentation layout will result in a human friendly errors.  These research shows the significance of improving the language in a compiler error message. Most principles and suggestions are followed in \chameleon{} in concustructing error statements. However they were not targeting type errors but compiler error (some even include runtime error) in general. The nuances of type errors, such as alternative typings were not taken into account.

@@

Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time. Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations. Provide opportunities to run consecutive command to query the information about code or executing command based on the existing code.
HazelTutor is a interactive filling type holes by suggesting template expressions (strategies as called by the authors). It also provide a cursor based type inspector that allows programmers to query the type of parts of the program.
Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y". It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.  These debugging tools are great motivations for the development of \chameleon{}. However they focus on different aspect of debugging. Java Whyline mainly tackle the problem of unintended runtime behavior while HazelTutor specialized in interactive actions supporting type holes.

>>>>

\section{Related Work}
%%%  Todo: Discussing "Practical SMT-based Type Error Localization"
Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process. Many studied [Citation here] the approach of finding all locations that contribute to a type error. Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.

One direction to approach type error debugging is by suggesting solutions to fix the type error. Sheng Chen's works \cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check. While this is a promising ground, the work is not evaluated with human participants. \cite{lerner_searching_2007} studied an approach of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed. This approach allows the algorithm to suggest syntax changes that could fix the type error. Benefiting from not needing a type system implementation and being language agnostic, the underlying idea was implemented and evaluated in two different languages with ten university students. While suggesting syntax changes is very intuitive in practice, this approach has limitations for teaching and learning purposes, as it is not able to explain why the type errors occur and why fixes are suggested. The lack of explanation is due to the inquiry into the underlying type-check being a binary state. Deep type-level analysis (typing rules and language features) was not communicated to the inference algorithm.

Another related approach is through ECEM (Enhanced compiler error message). Through a series of mixed-method studies, Prather showed \cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors. Decafe \cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version. In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices. Berik proposed a framework \cite{barik_how_2018} for constructing compiler error messages based on argumentation theory. In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.  These works show the significance of improving the language in the compiler error messages. Most principles and suggestions are followed in \chameleon{} in constructing error statements. However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general. The nuances of type errors, such as alternative typing, were not considered.

@@

Debugging using a GUI Interactive Development Environment has been the standard practice for a long time. Compared to a command line-based interface, a graphic user interface provides programming tools that can show information hierarchy, provide immediate feedback to changes, and display complex visualizations. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.  These debugging tools are important motivations for developing \chameleon{}. However, they focus on different contexts of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in interactive actions supporting type holes.
